Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-15 appendix.tex
more explanation of a-primes

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das:todo
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
Call the upper bound
$$
C_0(x^*, y^*) := \max\left(C^{1}_{\tau_1,\tau_1}(x^*, y^*), C^{2}_{\tau_1,\tau_1}(x^*, y^*), C^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
$$

>>>>

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The previous upper bound is for the ancestral state split $\{1\}$.
For example, for the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

We now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
Call the upper bound
$$
C_0(x^*, y^*) := \max\left(C^{1}_{\tau_1,\tau_1}(x^*, y^*), C^{2}_{\tau_1,\tau_1}(x^*, y^*), C^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
$$
